---
title: "Chapter 1 Visualization"
subtitle: "In-Class Exercises"
toc: true
format:
  html:
    self-contained: true
execute:
  warning: false
  eval: false
---

## Setup

First copy and past the following to load the **tidyverse** packages.

```{r}
library(tidyverse)
```

This will always be part of our setup.

## Palmer Penguins exercises

Use the Palmer Penguins data set introduced in the text to complete the following tasks.

1.  Load the package containing the `penguins` data set:

```{r}
library(palmerpenguins)
```

2.  Make a bar plot of `species` of `penguins`, where you assign `species` to the `y` aesthetic. How is this plot different?

3.  How are the following two plots different? Which aesthetic, `color` or `fill`, is more useful for changing the color of bars?

    ```{r}
    ggplot(penguins, aes(x = species)) +
      geom_bar(color = "red")

    ggplot(penguins, aes(x = species)) +
      geom_bar(fill = "red")
    ```

4.  Use the `penguins` data to determine what the `bins` argument in \`geom_histogram() does.

5.  Make a scatterplot of `bill_depth_mm` vs. `bill_length_mm` and color the points by `species`. What does adding coloring by species reveal about the relationship between these two variables? What about faceting by `species`?

6.  Why does the following yield two separate legends? How would you fix it to combine the two legends?

    ```{r}
    ggplot(
      data = penguins,
      mapping = aes(
        x = bill_length_mm, y = bill_depth_mm, 
        color = species, shape = species
      )
    ) +
      geom_point() +
      labs(color = "Species")
    ```

7.  Create the two following stacked bar plots. Which question can you answer with the first one? Which question can you answer with the second one?

    ```{r}
    ggplot(penguins, aes(x = island, fill = species)) +
      geom_bar(position = "fill")
    ggplot(penguins, aes(x = species, fill = island)) +
      geom_bar(position = "fill")
    ```

## Diamonds data set exercises

1.  The **diamonds** data is a rather large data set that is a bult-in part of the **tidyverse**. Enter the `?diamonds` command at the console to learn about the number of cases and variables. You can also enter `View(diamonds)` and `glimpse(dimaonds)` at the console to get a glimpse at the data.

2.  Make a histogram of the `carat` variable . Experiment with different binwidths. What binwidth reveals the most interesting patterns?
