---
title: "Introduction to String Manipulation in R"  
subtitle: "Chapter 14"
author: "Your Name Here"
output: 
  github_document: 
    toc: true
    toc_depth: 2
---


Many data sets have character strings in them. For example, in a file of tweets from Twitter (which are basically just strings of characters), perhaps you want to search for occurrences of a certain word or twitter handle. Or a character variable in a data set might be location with a city and state abbreviation, and you want to extract those observations with location containing "NY."


In this tutorial, you will learn how to manipulate text data using the package `stringr` and how to match patterns using regular expressions. The package `stringr` Some of the most commonly used commands include:


Command | Description
--------|---------------------------------------------------------
`str_sub` | Extract substring from a given start to end position
`str_detect` |Detect presence/absence of first occurrence of substring 
`str_locate` | Give position (start, end) of first occurrence of substring
`str_locate_all`|Give positions of all occurrences of a substring 
`str_replace` | Replace one substring with another



###1. Extracting and locating substrings


First, we introduce some basic commands from `stringr`.


The `str_sub` command extracts substrings from a string (that is, a sequence of characters) given the starting and ending position. For instance, to extract the characters in the second through fourth position or each string in fruits:


```{r}
library(stringr)
fruits <- c("apple", "pineapple", "Pear", "orange", "peach", "banana")
str_sub(fruits, 2, 4)
```


**Question 1.** What are the characters in the first through third position of each string in fruits?  

```{r}

```


**Question 2.** What happens if you enter a position number greater than the number of characters in the string?  Do this quietly and listen to hear whether R complains:)  

```{r}

```


**Question 3.** What happens if you enter just a single number? What if it's a negative number?  

```{r}

```


The `str_detect` command checks to see if any instance of a pattern occurs in a string.


```{r}
str_detect(fruits, "p")  #any occurrence of 'p'?
```


**Question 4:** Is pattern matching case-sensitive?  You should be able to tell by studying the logical results in the previous example. 


To locate the position of a pattern within a string, use `str_locate`:


```{r}

```

**Question 5:** Which fruits contain "an"?  

```{r}

```



**Question 6:** If there is more than one occurrence of the string, which occurence(s) is/are returned? (Hint: If you're having trouble with this, have a banana.)


To find all instances of "an" within each string:


```{r}

```

**Question 7:** What kind of object is returned by the `str_locate_all()` command immediately above?  

**Answer** 

The command `str_locate_all` returns a list. 

```{r}
out <- str_locate_all(fruits, "an")
data.class(out)
out[[6]]
unlist(out)
length(unlist(out))/2    #total number of times "an" occurs in vector fruits
```

The above example makes use of the fact that `str_locate_all()` returns a list. The command
`str_count()` provides an easier and more direct  way
to count the number of times "an" occurs in the names of the fruits:


```{r}
str_count(fruits, "an")
sum(str_count(fruits, "an"))
```


**Question 8:** Following the example immediately above, use `str_detect()` or `str_count()` command to output the 
number of these 6 words that contain the string "an". 

```{r}

```


###2. Regular expressions


Now suppose we want to detect or locate words that begin with "p" or end in "e," or match a more complex criteria. 
A **regular expression** is a sequence of characters that define a string pattern. It's a very powerful and flexible
albeit at times cryptic way to describe a text pattern of interest. Using the command `str_view()` and `str_view_all()` will be helpful for identifying what is actually being matched by each string detection. 

Let's detect strings that begin with either "p" or "P". The metacharacter "^" is used to indicate the beginning of the string, and "[Pp]" is used to indicate "P" or "p".


```{r}
str_view(fruits, "^[Pp]")
```




Similarly, the metacharacter "$" is used to signify the end of a string.


```{r}
str_view(fruits, "e$" )   #end in 'e'
```


The following are other metacharacters that have special meanings and so are reserved:

`* \ + $ { }  [ ]   ^ ? .`

For instance, a period matches any single character: 

gr.y matches gray, grey, gr9y, grEy, etc.


and * indicates 0 or more instances of the preceding character:

xy*z matches xz, xyz, xyyz, xyyyz, xyyyyz, etc.  


To detect the letter "a" followed by 0 or more occurrences of "p", type:


```{r}
str_view(fruits, "ap*")     
str_view_all(fruits, "ap*")
```


Compare this to


```{r}
str_view(fruits, "ap+")
```


The "+" after the "p" indicates that we want one or more occurrences of "p." 


Here is a more complex pattern:


```{r}
str_view(fruits, "^a(.*)e$") 
```


The anchors `^` and `$` are used to indicate we want strings that begin with the letter a and end with e. The `(.*)` indicates that we want to match 0 or more occurrences of any character. In particular, parentheses can be used to group parts of the pattern for readability.


###3 Example


Suppose we want to extract information on 10 digit United States phone numbers from a text file.


```{r}
a1 <- "Home: 507-645-5489"
a2 <- "Cell: 219.917.9871"
a3 <- "My work phone is 507-202-2332"
a4 <- "I don't have a phone"
info <- c(a1, a2, a3, a4)
info
```


We will now extract just the phone numbers from this string.


The area code must start with a 2 or higher so we use brackets again to indicate a range: [2-9]. The next two digits can be between 0 and 9, so we write [0-9]{2}. For the separator, we use [-.] to indicate either a dash or a period. The complete regular expression is given below:


```{r}
phone <- "([2-9][0-9]{2})[-.]([0-9]{3})[-.]([0-9]{4})"
str_view(info, phone)
```



Again, `str_view` just shows visually the presence or absence of the pattern in question. Instead, let's see how we could extract the phone numbers:


```{r}
str_extract(info, phone)
```


Let's anonymize the phone-numbers!


```{r}
str_replace(info, phone, "XXX-XXX-XXXX")
```

Or better, just anonymize the last 4 digits: 

```{r}
last4 <- "[0-9]{4}$" #Create the regexp for last 4 numbers
str_view(info, last4)
str_replace(info, last4, "XXXX")
```

**Question 9:** Can you use `str_replace_all` to replace all the periods by dash characters?  

```{r}

```




**Remarks** 

1. As we noted above, certain characters are reserved. If we want to actually reference them in a regular expression, either put them within a bracket, or use a double forward slash.


```{r}
str_locate(info, "[.]")  #find first instance of period
str_locate(info, "\\.")  #same
str_locate(info, ".")    #first instance of any character
```


2. Metacharacters have different meanings within brackets.  

For example, the caret character "^" indicates the starting point outside brackets but negation inside.


```{r}
str_detect(fruits, "^[Pp]")  #starts with 'P' or 'p'
str_detect(fruits, "[^Pp]")  #any character except 'P' or 'p'
str_detect(fruits, "^[^Pp]") #start with any character except 'P' or 'p'
```


3. See the  summary of regular expressions posted on Moodle under Chapter 14 Strings. This is a much more manageable (but still extensive!) set of regexp expressions as an alternative the the RStudio Cheat Sheet.


###4. Matching brackets or html tags


In many cases, you may want to match brackets such as [8] or html tags such as `<table>`.


```{r}
out <- c("abc[8]", "abc[9][20]", "abc[9]def[10][7]", "abc[]")
out

```


In order to better understand what regular expressions are matching here, we will replace pieces of the above strings with the character "X".


To match the left bracket, we write `\\[`. Next we want to match 0 or more occurrences of any character except the right bracket so we need `[^]]*` (remember what the caret does inside brackets?). Finally, to match the right bracket `\\]`. The parentheses are used to group parts of the expression and improve readability. The `str_replace_all()` command shows how the matched text could be replaced.


```{r}
str_view_all(out,  "\\[([^]]*)\\]")

str_replace_all(out, "\\[([^]]*)\\]", "X")
```


Compare this to


```{r}
str_view_all(out, "\\[(.*)\\]")

str_replace_all(out, "\\[(.*)\\]", "X")
```


In this case, we match the first left bracket (indicated by the `\\[`), followed by 0 or more instances of *any* character (the (.*) portion), which could be a right bracket until the final right bracket `\\]`. 


### On your own

1. Create a vector `veggies` containing "carrot", "bean", "peas", "cabbage", "scallion", "asparagus".

```{r}

```


+ Find those strings that contain the pattern "ea".

```{r}

```


+ Find those strings that end in "s".

```{r}

```


+ Find those strings that contain at least two "a"'s.

```{r}

```


+ Find those strings that begin with any letter except "c".

```{r}

```

+ Find the starting and ending position of the pattern "ca" in each string.

```{r}

```


2. The regular expression `"^[Ss](.*)(t+)(.+)(t+)"` matches "scuttlebutt", "Stetson", and "Scattter", but not "Scatter." Why?

```{r}

```


### Resources

* [http://www.regular-expressions.info/rlanguage.html](Regular expressions with R)