---
title: "Exploratory Data Analysis: Histograms, Frequency Polygons and More" 
author: "Your Name"
date: "`r Sys.Date()`"
output: 
  github_document: 
    toc: true
    toc_depth: 2
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(NHANES)
```



**1.** Create a histogram for the `Age` variable from the NHANES (National Health and Nutrition Examination Study ) study. What features do you observe? Is there anything particularly problematic? Try changing the binwidth to see if that affects your conclusions.  Another option to try is `breaks =` instead of bindwidth. What do you think `breaks = seq(0, 100, 25)` is doing? Try to use something more sensible than that. 

**Answer:**  

```{r}

```



**2.** Modify your previous chart so that height now represents the percent in the corresponding interval. Adjust the binwidth so that you can estimate fairly accurately the percent between the ages of 20 and 30. State your estimate. 

**Answer:**  

```{r}

```



**3.** Create separate frequency polygons for the variable `BPSysAve` for each category of the variable `BMI_WHO`. All the frequency polygons should appear on the same graph (not faceted). What are your observations? 

**Answer:**  


```{r}

```



**4.** In the previous question, did you notice a substantial number of NA values for the `BMI_WHO` variable? Do some digging and characterize who the participants are who were missing data on this variable. What do they have in common? Note: Don't just use `View()` but instead use the data wrangling or graphical tools we have learned. 

**Answer:**


```{r}

```

