---
title: Data Transformation Part 1 Practice
author: "Intro to Data Science"
date: "`r Sys.Date()`"
output: 
  github_document: 
    fig_height: 4
    fig_width: 9
    toc: true
    toc_depth: 2
---


```{r, setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(nycflights13)  
```

**Note:** Never put the backwards tick marks in the body of the R chunks.   I use them in the text here such as in `filter()`below  so that the commands names are clearer after you knit. 

# `filter()`


**`filter()`** picks observations by their values 

## Flights data examples


```{r}
flights  # note the flights is a tibble - a data.frame that has been tweaked
# use View(flights) to see entire data set
```

1) How do the variable types `int` and `dbl` differ?  What is `TRUE + TRUE`?

2) Find all the day and time variables. How are they stored?  


```{r}
filter(flights, month == 1, day == 1) 
(jan1 <- filter(flights, month == 1, day == 1))
```

3) How do the 2 `filter` lines above differ?


```{r}
filter(flights, month == 1 | month == 2) 
filter(flights, month == 1 | 2) 
```

4) Are the 2 filter commands above equivalent? If not, which do you prefer? What is the other one doing?  




5) Create 3 new data frames that include (a) flights going to Chicago Midway or O'Hare airport, (b) flights that were on time (not delayed), and (c) flights going to Chicago Midway or O'Hare airport that were on time.  What percentage of flights going to either of these two airports (from NYC) were on time in 2013?  

```{r}

```


### `filter()` examples:  

```{r}
# all flights in the spring: March, April, or May
filter(flights, month == 3 | month == 4 | month == 5)
filter(flights, month %in% c(3, 4, 5)) 

#  find all flights that left no more than a half hour after scheduled time
filter(flights, dep_delay <= 30)

#  find all flights that left no more than a half hour after scheduled time and arrived early
filter(flights, dep_delay <= 30 , arr_delay < 0)  

# two ways to find all flights not delayed by more than a half hour at either end
filter(flights, arr_delay <= 30, dep_delay <= 30)
filter(flights, !(arr_delay > 30 | dep_delay > 30))
```


6) Find the flights that were more than 500 miles long and arrived early.  


7) Find the flights that departed after midnight but no later than 7am. 

8) Find the flights that departed between midnight and 8am inclusive. 


9) What would happen if you tried `filter(flights, month == 3, month == 4)`?

10) How about `filter(flights, month = 3)`?  

11) Why does `filter(flights, month == (3 | 4))` give us flights in January?!?


12) What does this command do? `filter(flights, dest = DTW)`  


13) What proportion of Delta flights into DTW were delayed?


14) How many flights were missing departure times? (Hint: Use the `is.na()` function.)



15) Missing values (NAs) are a part of life whenever we are dealing with real data. Construct a general rule for how NAs are handled in R.  Consider examples like: `NA + 5`, `NA > 2`, `NA ^ 0`, `NA | TRUE`, `FALSE & NA`, `NA*0`.


# `arrange()` 

**`arrange()`** to reorder rows of data

## More flights data examples

```{r}
# order by year, then month within year, then day within month
#   2nd column breaks ties in the 1st, etc.
arrange(flights, year, month, day)

# get earliest departure time
arrange(flights, dep_time) 
arrange(flights, desc(dep_time))
```

16) What time was the earliest departure? 



17)  How are `NA` values sorted? This command displays the last few values and might be helpful: `tail(arrange(flights, desc(dep_time)))`  




18) Find the top 10 longest flights.  Where did they all go?  




# `select()` 

Use **`select()`** to pick variables

## Human Freedom Index Data examples

```{r}
hfi_data <- read_csv("data/hfi_cc_2018.csv")
```

Use `View(hfi_data)` to browse the data and refer to [this Kaggle link](https://www.kaggle.com/gsutters/the-human-freedom-index) for background information.


1) Select variables by name:

```{r}
# select certain variables by name
select(hfi_data, year: region)
select(hfi_data, year, countries, region, pf_score, ef_score, hf_score)

# By column number : NOT recommended in general b/c of reproducibility problems
select(hfi_data, 1:4 )

# select all columns except those between year and day
select(hfi_data, -(year:region))
```

2) Use `select()` to reorder

```{r}
# use select to put variables of most interest in front
select(hfi_data, year, countries, region, pf_score, ef_score, hf_score, everything())

# keeps all variables not explicitly mentioned - variant of select()
rename(hfi_data, country = countries)
```

3) Use the `select()` helper functions to pick variables  or groups of variables with similar names or distinguishing strings:

a) Select key variables and all personal freedom variables:  

```{r}
select(hfi_data, year, countries, region, starts_with("pf"))
```


b)  Select  all rank variables


```{r}
select(hfi_data, year, countries, region, ends_with("rank"))
```

c)  Select all regulation variables

```{r}
select(hfi_data, year, countries, region, contains("regulation"))
```

d)  Select all regulation variables

```{r}
# Select all gender- or sex-related  variables: either of these will work
select(hfi_data, year, countries, region, contains("sex"), contains("gender"))

select(hfi_data, year, countries, region, matches("sex|gender"))
```

4) Select all the variables that:  

* Contain references to `"tax"  

* Deal with "labor" freedoms  

* Contain references to "identity"  

* Deal with security and safety issues 

* In the first 4 or last 4 columns.


# `mutate()` 

Use **`mutate()`** to create new variables from old  

## More `flights` data examples

Because most of the variables in the `hfi_data` dataset are recorded on a 10-point scale, they don't lend themselves to much in the way of transformation. Instead, we return to the `flights` data to illustrate various ways of creating new variables.  

We first follow the text example to create a smaller data set of make it easier to see what's going on:

```{r}
flights_sml <- select(flights, year:day, ends_with("delay"), distance,
air_time )
```

Use `flights_sml` for the following `mutate()` examples:

1) Create  two new variables:

```{r}
mutate(flights_sml,
       gain = dep_delay - arr_delay, 
       speed = distance / air_time * 60
)
```

* Why not use `arr_delay - dep_delay`? 

* Why multiply by 60 in the speed calculation?  

2) Use `mutate()` to create a logical variable that indicates whether or not the flight was delayed in arriving?  


3) What's the difference between these two commands? 

```{r}
mutate(flights_sml, air_time_hrs = air_time/60)
```


```{r}
transmute(flights_sml, air_time_hrs = air_time/60)
```

When might you prefer the second over the first?  

4)  What is this doing? Explain the `%/%` and `%%` operators. 

```{r}
transmute(flights, dep_time,
          hour = dep_time %/% 100, 
          minute = dep_time %% 100
)
```


5) Use the data set `flights_sml2` and modulo arithmetic to convert actual departure time to departure time in minutes after midnight.  

```{r}
flights_sml2 <- select(flights, year:day, ends_with("time") )
```


```{r}
df <- mutate(flights_sml2,
       dep_hrs_after_mid = (dep_time %% 2400) %/% 100,
       dep_mins_aft_mid = 60 * dep_hrs_after_mid  + dep_time %% 100)
```



6) Find the airline and destination and date of the 5 most delayed (in arrival) flights using a ranking function. How do you want to handle ties? Carefully read the documentation for `min_rank()`. Hint: Use `select()` to show the variables of most interest first.  






