---
title: "Chapter 11: Data Import Practice"
output: 
  github_document: 
    toc: true
    toc_depth: 2
---


```{r, setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

Because of the technical nature of this section, we'll follow the text very closely and try to understand the tools and tricks for correctly importing problematic data. Few datasets will have all the issues raised in this section but many will have some.  

### Some Basics of Reading with `read_csv()`

Start by creating an inline `csv` file. This is useful for experimenting with `readr` and for creating reproducible examples to share with others. Running the lines below creates a **tibble** with 2 rows and 3 columns:

```{r}
read_csv("a,b,c
1,2,3
4,5,6")
```

1. You can use `skip = n` to skip the first n lines of metadata:  

```{r}
read_csv("The first line of metadata 
The second line of metadata
x,y,z
1,2,3", skip = 2)
```

Can also ignore all lines preceded by a specified comment character:  

```{r}
read_csv("# A comment I want to skip
x,y,z
# Another comment
1,2,3", comment = "#")
```

Question: What would have happened if we didn't specify the comment character? 


2. If no column names, let R generate them or specify them manually. 

```{r}
# R generates the names
read_csv("1,2,3\n 4,5,6", col_names = FALSE)
```
Note use of the newline specifier `\n`.

```{r}
# We specify the names 
read_csv("1,2,3\n4,5,6", col_names = c("x", "y", "z"))
```

3. Specify `na` values if non-standard NA values used:  

```{r}
read_csv("a,b,c\n1,2,.", na = ".")
```


Take comfort in these words from Hadley Wickham: 

> "This is all you need to know to read ~75% of CSV files that you'll encounter in practice. You can also easily adapt what you’ve learned to read tab separated  files with read_tsv() and  fixed width files with read_fwf(). To read in more challenging  files, you'll need to learn more about how **readr** parses each column, turning them into R vectors."

Practice Exercises:  

P1. What function would you use to read a file where fields were separated with the vertical line `|`?  

P2. Identify what is wrong with each of the following inline CSV file. What happens when you run the code?  (Try to spot the error before running the command.)

```{r}
#(a)
read_csv("a,b\n1,2,3\n4,5,6")
#(b)
read_csv("a,b,c\n1,2\n1,2,3,4") 
#(c)
read_csv("a,b\n\"1") 
#(d)
read_csv("a,b\n1,2\na,b") 
#(e)
read_csv("a;b\n1;3")
```


### Parsing a Vector

**parsing:** "the process of analyzing a string of symbols, either in natural language, computer languages or data structures, conforming to the rules of a formal grammar."

Parsing functions are building blocks of the `readr` functions:  

What are these doing? 

```{r}
str(parse_logical(c("TRUE", "FALSE", "NA")))
str(parse_integer(c("1", "2", "3")))
str(parse_date(c("2010-01-01", "1979-10-14"))) 
parse_integer(c("1", "231", ".", "456"), na = ".")
```

How to see if parsing has failed? 

```{r}
x <- parse_integer(c("123", "345", "abc", "123.45")) 
x 
# Use problems() to see what went wrong
problems(x)
```

These procedures help you see how to "fix" your specification (data?) so that R can correctly import it. 

### Parsing Numbers  

Key considerations in identifying and correctly parsing numbers: 

1. Decimal point is locale dependent (US vs Europe, etc)  

2. Surrounding characters: percent, dollar sign, etc.  

3. Grouping characters: comma in US, period in Europe  

```{r}
# parse_double and decimal point
parse_double("1.23")
parse_double("1,23", locale = locale(decimal_mark = ","))
```

```{r}
# parse_number and surrounding characters
parse_number("$100")
parse_number("20%")
parse_number("It cost $123.45")
```

```{r}
# parse_number and grouping characters
# Used in America
parse_number("$123,456,789")
# Used in many parts of Europe
parse_number("123.456.789", locale = locale(grouping_mark = "."))
```

### Parsing Dates, date-times, and times  

Three common parsers in R:  
* `parse_date()`:  expects something like: YYYY-MM-DD  
* `parse_time()`: expects something like: HH:MM::SS  
* `parse_datetime()`: expects an international standard ISO8601 date-time

Examples below:  

```{r}
parse_datetime("2010-10-01T0830")
parse_datetime("20101010")
```


```{r}
parse_date("2016-01-29")
```

```{r}
library(hms)
parse_time("01:10 am")
parse_time("20:10:01")
```

Much of the time your data will not fit the standard formats R expects and you'll need to specify custom formats using the date-time formats in Section 11.3.4.  

P3. Use the formats in Section 11.3.4 to generate the correct format string to parse each of the following dates and times:  

```{r}
d1 <- "January 1, 2010"
d2 <- "2015-Mar-07"
d3 <- "06-Jun-2017"
d4 <- c("August 19 (2015)", "July 1 (2015)") 
d5 <- "12/30/14" # Dec 30, 2014
t1 <- "1705"
t2 <- "11:15:10.12 PM"
```

### Parsing a File  

P4. The basic function used by **readr** to try to identify file types is `guess_parser()`. Try the examples provided and fix any that fail. 

```{r}
guess_parser("2010-10-01")
guess_parser("15:O1")
guess_parser(c("TRUE", "FALSER")) 
guess_parser(c("1", "5", "9"))
guess_parser(c("12,352,56l"))
```


Now let's try using one of R's challenge files to see what could go wrong:  

1. First pass:  

```{r}
challenge <- read_csv(readr_example("challenge.csv"))
```

Two printed outputs:   
    + Guess at column specification based on first 1000 rows  
    + First five parsing "failures"

2. Second pass:  

  + First column guess seems ok  
  + Override default guess at second column type to see what's going on

```{r}
challenge <- read_csv(
  readr_example("challenge.csv"),
  col_types = cols(
    x = col_double(),
    y = col_character() 
    )
  )
```

No errors? Let's take a look at what it read:  

```{r}
challenge
head(challenge)
tail(challenge)
```

3. Third pass:  

  + Fix the specification on the second column because we think it's a date when non-missing:  
 
```{r}
challenge <- read_csv(
  readr_example("challenge.csv"),
  col_types = cols(
    x = col_double(),
    y = col_date() 
    )
  )
``` 
  
Check again to make sure:  
  
```{r}
head(challenge)
tail(challenge)
```

Declare success!! 

If this doesn't seem to work, try changing the `n_max` value to be more than 1,000 or read line by line (`read_lines()`).

Note:  There's not always one correct answer. The right answer may depend on the end use (and the end user!). 