---
title: "Your First Quarto Document"
subtitle: "UN Voting Data Visualization"
author: "Your Name"
date: today
format: gfm
toc: true
editor: visual
---

## Goals

-   Get acquainted with the Quarto document look and feel
-   Practice with rendering a Quarto document to a **gfm** file (GitHub-flavored Markdown)
-   See more examples of the use of **ggplot2**
-   Preview the kinds of data transformations to come
-   Preview the use of the pipe (`|>`) for chaining functions

## Background

Let's take a look at the voting history of countries in the United Nations General Assembly. We will be using data from the **unvotes** package. Additionally, we will make use of the **tidyverse** and **lubridate** packages for the analysis, and the **DT** package for interactive display of tabular output.

```{r}
#| label: setup
#| include: false
library(unvotes)
library(tidyverse)
library(lubridate)
```

## Data

The unvotes package provides three data sets we can work with: `un_roll_calls`, `un_roll_call_issues`, and `un_votes`. Each of these datasets contains a variable called `rcid`, the roll call id, which can be used as a unique identifier to join them with each other.

-   The `un_votes` data set provides information on the voting history of the United Nations General Assembly. It contains one row for each country-vote pair.

```{r}
#| label: voew-un_votes
un_votes
```

-   The `un_roll_calls` dataset contains information on each roll call vote of the United Nations General Assembly.

```{r}
#| label: view-un_roll_calls
un_roll_calls
```

-   The `un_roll_call_issues` dataset contains (topic) classifications of roll call votes of the United Nations General Assembly. Many votes had no topic, and some have more than one.

```{r}
#| label: view-un_roll_call_issues
un_roll_call_issues
```

## Analysis

First, let's take a look at how often each country voted "Yes" on a resolution in each year. We'll visualize the results, so let's pick a few countries of interest first,

```{r}
#| label: country-list
country_list <- c("United States", "Russia", 
                  "Mexico", "Turkey")
```

and focus our analysis on them.

```{r}
#| label: plot-yearly-yes
#| warning: false
#| fig-width: 10
un_votes |>
  filter(country %in% country_list) |>
  inner_join(un_roll_calls, by = "rcid") |>
  group_by(year = year(date), country) |>
  summarize(
    votes = n(),
    percent_yes = mean(vote == "yes")
    ) |>
  ggplot(mapping = aes(x = year, y = percent_yes, color = country)) +
    geom_line() +
    labs(
      title = "Percentage of 'Yes' votes in the UN General Assembly",
      subtitle = "1946 to 2019",
      y = "% Yes",
      x = "Year",
      color = "Country"
    )
```

Next, let's create a visualization that displays how the voting record of the United States changed over time on a variety of issues, and compares it to another country. The other country we'll display is India.

```{r}
#| label: plot-yearly-yes-issue
#| warning: false
#| fig-width: 10
un_votes |>
  filter(country %in% c("United States", "India")) |>
  inner_join(un_roll_calls, by = "rcid") |>
  inner_join(un_roll_call_issues, by = "rcid") |>
  group_by(country, year = year(date), issue) |>
  summarize(
    votes = n(),
    percent_yes = mean(vote == "yes")
    ) |>
  filter(votes > 5) |>  # only use records where there are more than 5 votes
  ggplot(mapping = aes(x = year, y = percent_yes, color = country)) +
    geom_point() +
    geom_smooth(method = "loess", se = FALSE) +
    facet_wrap(~ issue) +
    labs(
      title = "Percentage of 'Yes' votes in the UN General Assembly",
      subtitle = "1946 to 2019",
      y = "% Yes",
      x = "Year",
      color = "Country"
    )
```

We can easily change which countries are being plotted by changing which countries the code above `filter`s for. Note that the country name should be spelled and capitalized exactly the same way as it appears in the data. See the [Appendix](#appendix) for a list of the countries in the data.

## Exercises

1.  Try modifying the code above substituting the names of other countries. (You can browse the **un_votes** data set to check the spelling.)
2.  Try modifying aesthetics of the graphs.
3.  What does `method = "loess"` mean?

Source of this activity: [*Data Science in a Box*](https://datasciencebox.org/).

## References

1.  David Robinson (2017). unvotes: United Nations General Assembly Voting Data. R package version 0.2.0. https://CRAN.R-project.org/package=unvotes.
2.  Erik Voeten "Data and Analyses of Voting in the UN General Assembly" Routledge Handbook of International Organization, edited by Bob Reinalda (published May 27, 2013).
3.  Much of the analysis has been modeled on the examples presented in the [unvotes package vignette](https://cran.r-project.org/web/packages/unvotes/vignettes/unvotes.html).
